---
title: "Designing LSVI-mapping in effect monitoring for the Programmatic Nitrogen Approach (PAS)"
date: today
published-title: Date
# csl: "`r if (curl::has_internet()) 'https://raw.githubusercontent.com/citation-style-language/styles/master/research-institute-for-nature-and-forest.csl' else system.file('research-institute-for-nature-and-forest.csl', package = 'INBOmd')`"
# bibliography: ../../../refs_mne_design.yaml
params:
  lang: nl
  rewrite_subcellpanels: FALSE
engine: knitr
format:
  html:
    embed-resources: true
    include-in-header: header_qmd.html
    theme: cosmo
    keep-md: true
    number-sections: true
    code-fold: true
    fig-width: 10
    fig-height: 5
    df-print: paged
    toc: true
    toc-location: left
    toc-depth: 4
    toc-expand: true
---

```{r setup, message = FALSE}
options(scipen = 999)
conflictRules("git2rdata", exclude = c("pull", "meta"))
conflictRules("tidyr", exclude = "extract")
conflictRules("knitr", exclude = "spin")
conflictRules("kableExtra", exclude = "group_rows")
lang <- params$lang
library(rprojroot)
projroot <- find_root(is_rstudio_project)
gitroot <- find_root(is_git_root)
datapath <- file.path(projroot, "data")
for (subdir in c(
  "binary/results",
  "text/raw",
  "text/results"
)) {
  path <- file.path(datapath, subdir)
  if (!dir.exists(path)) dir.create(path, recursive = TRUE)
}
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(git2rdata)
library(sf)
library(units)
library(terra)
library(n2khab)
library(ggplot2)
library(mapview)
source(file.path(projroot, "R/functions.R"))
opts_chunk$set(
  echo = TRUE,
  dpi = 300,
  fig.align = "center"
)
our_kbl <- function(df) {
  df %>% 
    kbl(table.attr = 'data-quarto-disable-processing="true"') %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "striped")
}
our_scroll_kbl <- function(df, width = NULL, height = NULL) {
  our_kbl(df) %>% 
    scroll_box(width = width, height = height)
}
```

## Settings and basic objects

Defining the regular expression for forest types.

```{r}
#| code-fold: show
forest_types_regex <- "^9|^2180|^rbbppm"
```

Setting the theoretical first year of the first monitoring cycle as defined for reporting, i.e. taking into account the required alignment of monitoring cycles with reporting years.

```{r}
#| code-fold: show
year_origin <- 2024L
```

Reading the reference checksum values.

```{r}
n2khab_data_checksums_reference <-
  read_vc(
    file = "n2khab_data_checksums_reference",
    root = file.path(datapath, "text/raw")
  ) %>%
  as_tibble()
```

Defining the data source versions to be used in this workflow.

```{r}
#| code-fold: show
versions_required <- c(
  sac = "sac_2013-01-18",
  habitatmap_terr = "habitatmap_terr_2025_v99_interim",
  GRTSmh_diffres = "GRTSmh_diffres_v1"
)
stopifnot(all(
  versions_required %in% unique(n2khab_data_checksums_reference$version_id)
))
```

Checking the existence of the correct files in the correct directories.

```{r}
verify_n2khab_data(n2khab_data_checksums_reference, versions_required)
```

Defining domains, including their geometry.

```{r}
sac <- read_admin_areas(dsn = "sac")
sac_subset_5 <- c(
  "BE2100015", # Kalmthoutse Heide
  "BE2100024", # Vennen, heiden en moerassen rond Turnhout
  "BE2200028", # De Maten
  "BE2200035", # Mechelse heide en vallei van de Ziepbeek
  "BE2200039" # Voerstreek
)
for (sac_i in sac_subset_5) {
  sac %>%
    filter(sac_code == sac_i) %>%
    st_combine() %>%
    .[[1]] %>%
    assign(str_c("sfg_", tolower(sac_i)), ., envir = .GlobalEnv)
}
domains <- tribble(
  ~domain, ~domain_name_nl, ~domain_name_en, ~geometry,
  "BE2100015", "SBZ-H BE2100015", "SAC BE2100015", sfg_be2100015,
  "BE2100024", "SBZ-H BE2100024", "SAC BE2100024", sfg_be2100024,
  "BE2200028", "SBZ-H BE2200028", "SAC BE2200028", sfg_be2200028,
  "BE2200035", "SBZ-H BE2200035", "SAC BE2200035", sfg_be2200035,
  "BE2200039", "SBZ-H BE2200039", "SAC BE2200039", sfg_be2200039
) %>%
  mutate(across(where(is.character), fct)) %>%
  st_as_sf(crs = 31370)
domains
```

Reading spatial data of terrestrial types.

```{r}
#| code-fold: show
hmt <- read_habitatmap_terr(
  keep_aq_types = FALSE,
  drop_7220 = FALSE,
  version = versions_required["habitatmap_terr"]
)
```

Reading GRTS raster cells at two specific resolutions.

```{r}
#| code-fold: show
main_level <- 4
cells <- read_GRTSmh_diffres(level = main_level)
subcells <- read_GRTSmh_diffres(level = main_level - 1)
```

## Subsetting objects

Only polygons with a terrestrial _habitat_ type are retained.
We also mark which polygons contain at least one 'open type' (as opposed to 'forest type').

```{r}
hmt_occ_hab <- 
  hmt$habitatmap_terr_types %>%
  filter(!str_detect(type, "^rbb")) %>% 
  mutate(
    polygon_has_open_type = !all(str_detect(type, forest_types_regex)),
    .by = polygon_id
  )
hmt_pol_hab <- 
  hmt$habitatmap_terr_polygons %>% 
  inner_join(
    hmt_occ_hab %>%
      distinct(
        polygon_id,
        has_open_type = polygon_has_open_type
      ),
    join_by(polygon_id),
    relationship = "one-to-one",
    unmatched = c("drop", "error")
  ) %>% 
  select(polygon_id, description_orig, has_open_type)
```

Spatially subsetting habitat polygons according to domains (five SACs with custom measures), i.e. all habitat polygons are retained that intersect the domains.

```{r}
hmt_pol_hab_sacsubset <- hmt_pol_hab[domains, ]
```

Inspecting result:

```{r}
#| paged.print: false
#| code-fold: show
hmt_pol_hab_sacsubset
```

```{r}
#| code-fold: show
table(hmt_pol_hab_sacsubset$has_open_type)
```

Spatially subsetting cells and subcells according to domains.
The (sub)cells are retained when they overlap the domain (the cell center does not have to overlap the domain).
We also include the cell address in the subcells object.

```{r}
cells_sacsubset_pre <-
  cells[vect(domains), , drop = FALSE] %>% 
  as.polygons(round = FALSE, aggregate = FALSE) %>% 
  st_as_sf() %>% 
  as_tibble() %>% 
  rename(cell_address = 1) %>% 
  mutate(cell_address = as.integer(cell_address)) %>% 
  st_as_sf()
# adding domain attribute
cells_sacsubset <- 
  cells_sacsubset_pre %>% 
  st_join(
    domains %>% select(domain),
    join = st_intersects,
    left = FALSE
  ) %>% 
  relocate(geometry, .after = last_col()) %>% 
  arrange(domain, cell_address)
# we assume that no cells are member of two domains at the same time (if not
# fulfilled we need to re-evaluate how we cope with domain-wise cell panel
# membership)
stopifnot(nrow(cells_sacsubset_pre) == nrow(cells_sacsubset))
subcells_sacsubset <-
  subcells[vect(domains), , drop = FALSE] %>% 
  as.polygons(round = FALSE, aggregate = FALSE) %>% 
  st_as_sf() %>% 
  as_tibble() %>% 
  rename(subcell_address = 1) %>% 
  mutate(subcell_address = as.integer(subcell_address)) %>% 
  st_as_sf() %>% 
  st_join(cells_sacsubset, join = st_within, left = FALSE) %>% 
  arrange(cell_address, subcell_address) %>% 
  relocate(geometry, .after = last_col())
```

Inspecting the results:

```{r}
#| paged.print: false
#| code-fold: show
cells_sacsubset
```

```{r}
#| paged.print: false
#| code-fold: show
subcells_sacsubset
```

Visualizing domains, cells and subcells.

```{r}
#| fig-height: 8
#| fig-width: 8
mvmap <- 
  mapview(
    subcells_sacsubset,
    map.types = "CartoDB.Positron",
    col.regions = "blue"
  ) +
  mapview(cells_sacsubset, col.regions = "red", lwd = 1.5, alpha.regions = 0.5) +
  mapview(
    domains,
    alpha.regions = 0,
    col.regions = "pink",
    color = "pink",
    lwd = 2
  )
mvmap@map %>% 
  inbospatial::add_wms_be_cartoweb("topo_grey")
```

## Habitat polygons with (sub)cell attribute

To join subcell and cell addresses to the habitat polygons in a one-to-one way, we will use the centroids of the habitat polygons.

First creating these centroids.

```{r}
hmt_pol_hab_sacsubset_centroid <- 
  st_intersection(
    hmt_pol_hab_sacsubset %>% st_sf(agr = "identity"),
    domains %>% st_geometry()
  ) %>% 
  st_sf(agr = "identity") %>% 
  st_centroid()
```

```{r}
stopifnot(nrow(hmt_pol_hab_sacsubset) == nrow(hmt_pol_hab_sacsubset_centroid))
```

```{r}
#| include: false
if (FALSE) {
  cells_sacsubset_hmt <-
    cells_sacsubset[hmt_pol_hab_sacsubset_centroid, ]
  # all subcells belonging to cells_sacsubset_hmt
  subcells_sacsubset_hmt <-
    subcells_sacsubset %>% 
    semi_join(
      cells_sacsubset_hmt %>% st_drop_geometry(),
      join_by(cell_address)
    )
}
```

```{r}
#| include: false
if (FALSE) {
  mapview::mapview(domains) +
    cells_sacsubset_hmt +
    subcells_sacsubset_hmt +
    hmt_pol_hab_sacsubset_centroid
}
```

Joining subcell and cell addresses to the habitat polygons.

```{r}
hmt_pol_hab_sacsubset_subcell <- 
  hmt_pol_hab_sacsubset_centroid %>% 
  st_join(subcells_sacsubset, left = FALSE) %>% 
  st_drop_geometry() %>% 
  inner_join(
    hmt_pol_hab_sacsubset,
    .,
    join_by(polygon_id, description_orig, has_open_type),
    relationship = "one-to-one",
    unmatched = "error"
  ) %>% 
  relocate(geom, .after = last_col())
```

Inspecting result:

```{r}
#| paged.print: false
#| code-fold: show
hmt_pol_hab_sacsubset_subcell
```


## Implement membership design

The panel membership will be defined at the level of cells and subcells, using their (reverse) GRTS address order.

For panel memberships, we start from the `cells_sacsubset` and `subcells_sacsubset` objects without taking into account habitat polygon distribution, to make these membership more stable through time, i.e. not sensitive to changes in the habitatmap.
This approach is expected to still lead to a relatively even spread of the habitat polygons over panels, which we will check.

Note that for subcells we use _subpanels_, i.e. a numbering to be interpreted within the cell panel.

We always recalculate the cell panel membership.
If the domains ever change, this may cause a few shifts but these should still be relatively small given the large number of cells compared to the number of panels.
Still, we will protect against effects of domain additions or removals by calculating panel membership domain-wise, not across domains.

```{r}
set.seed(20260908) # not clear whether this may have effect
cells_sacsubset_panel <- 
  cells_sacsubset %>% 
  st_drop_geometry() %>% 
  mutate(
    panel_cell = cell_address %>% 
      cut_number(n = 6) %>% 
      as.integer(),
    .by = domain
  )
```

Distribution of cells over panels:

```{r}
#| code-fold: show
table(cells_sacsubset_panel$panel_cell)
```

We have calculated subcell membership of subpanels in the past and saved it on disk.
We recalculate the subpanel membership (and optionally rewrite it) to test for equality with the version on disk.

```{r}
set.seed(20260908)
subcells_sacsubset_subpanel_recalc <- 
  subcells_sacsubset %>% 
  st_drop_geometry() %>% 
  inner_join(
    cells_sacsubset_panel,
    join_by(cell_address, domain),
    relationship = "many-to-one",
    unmatched = "error"
  ) %>% 
  mutate(
    subpanel_subcell = divide_subcell_addresses_in_two(subcell_address),
    .by = cell_address
  )
```

```{r}
if (
  params$rewrite_subcellpanels ||
  !file.exists(file.path(
    datapath,
    "text/results/subcells_sacsubset_subpanel.yml"
  ))
) {
  subcells_sacsubset_subpanel_recalc %>% 
    write_vc(
      file = "subcells_sacsubset_subpanel",
      root = file.path(datapath, "text/results"),
      digits = 15,
      optimize = TRUE,
      strict = FALSE,
      sorting = c("cell_address", "subcell_address")
    ) %>% 
    convert_line_endings(root = file.path(datapath, "text/results"))
}
```

Using the subcell subpanel membership that has been saved before, and test that it is the same.

```{r}
subcells_sacsubset_subpanel <- 
  read_vc(
    file = "subcells_sacsubset_subpanel",
    root = file.path(datapath, "text/results")
  ) %>% 
  as_tibble()
# if this test fails, manual intervention is needed to decide on changes
stopifnot(all.equal(
  subcells_sacsubset_subpanel,
  subcells_sacsubset_subpanel_recalc,
  check.attributes = FALSE
))
```

## Define revisit design

We define the revisit design as an assignment of calendar years to panels and subpanels for a period of 24 years.

The revisit design is different between polygons that have at least one open type and forest polygons:

- when open types are present, a cycle duration of 6 years is used in a serially alternating pattern [1-5];
- when a polygon is forest-only, a cycle duration of 12 years is used in a serially alternating pattern [1-11], implemented by creating 2 _subpanels_ within each panel for open types.
In this way the panel membership of open types is taken into account.

Generating the panel calendar.

```{r}
panel_calendar <- 
  subcells_sacsubset_subpanel %>% 
  expand(
    panel_cell = full_seq(panel_cell, 1L),
    subpanel_subcell = full_seq(subpanel_subcell, 1L)
  ) %>% 
  arrange(subpanel_subcell, panel_cell) %>% 
  bind_rows(., .) %>% 
  bind_cols(year_lsvi = year_origin + seq_len(nrow(.)) - 1L)
```

Inspecting the result:

```{r}
#| paged.print: false
#| code-fold: show
panel_calendar %>% print(n = Inf)
```

## Generate calendars for cells, subcells and polygons

We generate the cell calendar (open types), the subcell calendar (forest types) and polygon calendar by joining the panel years to the (sub)panels.
Only in the polygon calendar it is possible to integrate both revisit designs without spatial overlaps.

```{r}
cells_calendar <- 
  cells_sacsubset_panel %>% 
  inner_join(
    panel_calendar %>% 
      select(panel_cell, year_lsvi),
    join_by(panel_cell),
    relationship = "many-to-many",
    unmatched = "error"
  )
```

Structure of the cell calendar (open types):

```{r}
#| code-fold: show
glimpse(cells_calendar)
```

```{r}
subcells_calendar <- 
  subcells_sacsubset_subpanel %>% 
  inner_join(
    panel_calendar,
    join_by(panel_cell, subpanel_subcell),
    relationship = "many-to-many",
    unmatched = "error"
  )
```

Structure of the subcell calendar (forest types):

```{r}
#| code-fold: show
glimpse(subcells_calendar)
```

In the polygon calendar, we set the subcell address and subcell subpanel as missing if the polygon has at least one open type, to avoid confusion: subcells are not relevant in the revisit design of those polygons.

```{r}
polygons_calendar <- 
  hmt_pol_hab_sacsubset_subcell %>% 
  st_drop_geometry() %>% 
  inner_join(
    cells_calendar,
    join_by(cell_address, domain),
    relationship = "many-to-many",
    unmatched = c("error", "drop")
  ) %>% 
  left_join(
    subcells_calendar,
    join_by(
      subcell_address,
      cell_address,
      domain,
      panel_cell,
      # matching years: (forest) subcell revisit years match (open) cell revisit
      # years
      year_lsvi
    ),
    # multiple polygons can have their centroid in the same subcell
    relationship = "many-to-one",
    unmatched = "drop"
  ) %>% 
  # keep subcell_address and subpanel_subcell only for forest-only polygons
  mutate(across(
    c(subcell_address, subpanel_subcell),
    \(x) ifelse(has_open_type, NA_integer_, x)
  )) %>% 
  # drop the empty subcell revisits in case of forest-only polygons
  filter_out(!has_open_type & is.na(subpanel_subcell)) %>% 
  relocate(
    c(
      year_lsvi,
      domain,
      cell_address,
      panel_cell,
      subcell_address,
      subpanel_subcell
    ),
    .after = has_open_type
  ) %>% 
  arrange(year_lsvi, domain, cell_address, subcell_address)
```

Structure of the polygon calendar (all types):

```{r}
#| code-fold: show
glimpse(polygons_calendar)
```


## Derive applied, spatial objects

We also derive geospatial objects that flatten the years in a single row.
These objects are stripped from background information to avoid confusion in applications.

```{r}
opentypes_calendar_lsvi_pas <-
  cells_sacsubset %>% 
  inner_join(
    cells_calendar %>% 
      summarise(
        years_lsvi = str_flatten(sort(year_lsvi), collapse = "|") %>% factor(),
        .by = c(cell_address, domain)
      ),
    join_by(cell_address, domain),
    relationship = "one-to-one",
    unmatched = "error"
  ) %>%
  relocate(geometry, .after = last_col()) %>% 
  arrange(domain, years_lsvi, cell_address)
```

Inspecting geospatial form of the cell calendar (open types):

```{r}
#| paged.print: false
#| code-fold: show
opentypes_calendar_lsvi_pas
```

```{r}
foresttypes_calendar_lsvi_pas <-
  subcells_sacsubset %>% 
  select(-cell_address) %>% 
  inner_join(
    subcells_calendar %>% 
      summarise(
        years_lsvi = str_flatten(sort(year_lsvi), collapse = "|") %>% factor(),
        .by = c(subcell_address, domain)
      ),
    join_by(subcell_address, domain),
    relationship = "one-to-one",
    unmatched = "error"
  ) %>%
  relocate(geometry, .after = last_col()) %>% 
  arrange(domain, years_lsvi, subcell_address)
```

Inspecting geospatial form of the subcell calendar (forest types):

```{r}
#| paged.print: false
#| code-fold: show
foresttypes_calendar_lsvi_pas
```

```{r}
polygons_calendar_lsvi_pas <-
  hmt_pol_hab_sacsubset %>% 
  inner_join(
    polygons_calendar %>% 
      summarise(
        years_lsvi = str_flatten(sort(year_lsvi), collapse = "|") %>% factor(),
        .by = c(polygon_id, description_orig, has_open_type, domain)
      ),
    join_by(polygon_id, description_orig, has_open_type),
    relationship = "one-to-one",
    unmatched = "error"
  ) %>%
  relocate(geom, .after = last_col()) %>% 
  arrange(domain, years_lsvi, polygon_id)
```

Inspecting geospatial form of the polygon calendar (all types):

```{r}
#| paged.print: false
#| code-fold: show
polygons_calendar_lsvi_pas
```


## Explore results

```{r}
plot_category_areas <- function(df, response_var, fill = NULL) {
  df %>% 
  mutate({{response_var}} := round({{response_var}}, 2)) %>% 
  ggplot(aes(
    x = year_lsvi,
    y = {{response_var}},
    fill = {{fill}},
    label = {{response_var}}
  )) +
  geom_col() +
  scale_x_continuous(breaks = seq(2024, 2047, 1)) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.4),
    legend.position = "bottom"
  )
}
```

### Visualizing revisit design: total cell and subcell numbers per year

```{r}
cell_count_yr <- 
  cells_sacsubset %>% 
  select(cell_address) %>% 
  st_drop_geometry() %>% 
  inner_join(
    cells_calendar %>%
      select(domain, cell_address, year_lsvi),
    join_by(cell_address),
    relationship = "one-to-many",
    unmatched = "error"
  ) %>% 
  count(domain, year_lsvi, name = "number_of_cells")
```

Total number of cells per year:

```{r}
cell_count_yr %>% 
  summarise(number_of_cells = sum(number_of_cells), .by = year_lsvi) %>% 
  our_kbl()
```

```{r}
cell_count_yr %>% 
  plot_category_areas(response_var = number_of_cells, fill = domain) +
  ggtitle("Total number of cells intersecting the domains, irrespective of the presence of habitat")
```

```{r}
subcell_count_yr <- 
  subcells_sacsubset %>% 
  select(subcell_address) %>% 
  st_drop_geometry() %>% 
  inner_join(
    subcells_calendar %>%
      select(domain, subcell_address, year_lsvi),
    join_by(subcell_address),
    relationship = "one-to-many",
    unmatched = "error"
  ) %>% 
  count(domain, year_lsvi, name = "number_of_subcells")
```

Total number of subcells per year:

```{r}
subcell_count_yr %>% 
  summarise(number_of_subcells = sum(number_of_subcells), .by = year_lsvi) %>% 
  our_kbl()
```

```{r}
subcell_count_yr %>% 
  plot_category_areas(response_var = number_of_subcells, fill = domain) +
  ggtitle("Total number of subcells intersecting the domains, irrespective of the presence of habitat")
```

### Visualizing LSVI surface areas by year

Total areas according to `has_open_type` and panel (as represented by groups of years).
The table rows represent mutually exclusive polygon sets.

```{r}
polygons_calendar_lsvi_pas %>% 
  mutate(polygon_area = st_area(.) %>% set_units("ha")) %>% 
  st_drop_geometry() %>% 
  summarise(
    surface_area = sum(polygon_area) %>% round(2),
    .by = c(has_open_type, years_lsvi)
  ) %>% 
  arrange(desc(has_open_type), years_lsvi) %>% 
  our_kbl()
```

The same information, visualized by year:

```{r}
surface_areas_yr <- 
  hmt_pol_hab_sacsubset %>% 
  select(polygon_id) %>% 
  mutate(polygon_area = st_area(.) %>% set_units("ha")) %>% 
  st_drop_geometry() %>% 
  inner_join(
    polygons_calendar %>%
      select(domain, polygon_id, has_open_type, year_lsvi),
    join_by(polygon_id),
    relationship = "one-to-many",
    unmatched = "error"
  ) %>% 
  mutate(polygon_category = ifelse(
    has_open_type,
    "contains_opentype",
    "forest_only"
  )) %>% 
  summarise(
    surface_area = sum(polygon_area),
    .by = c(domain, polygon_category, year_lsvi)
  )
```

```{r}
surface_areas_yr %>% 
  summarise(
    surface_area = sum(surface_area),
    .by = c(polygon_category, year_lsvi)
  ) %>% 
  plot_category_areas(response_var = surface_area, fill = polygon_category) +
  geom_text(size = 3, vjust = 2)
```

Distinguishing domains:

```{r}
surface_areas_yr %>% 
  summarise(
    surface_area = sum(surface_area),
    .by = c(domain, polygon_category, year_lsvi)
  ) %>% 
  plot_category_areas(response_var = surface_area, fill = polygon_category) +
  facet_wrap(~domain, scales = "free_y")
```

Total areas by year:

```{r}
surface_areas_yr_total <- 
  surface_areas_yr %>% 
  summarise(
    surface_area = sum(surface_area),
    .by = c(year_lsvi)
  ) %>% 
  mutate(surface_area = round(surface_area, 2)) %>% 
  arrange(year_lsvi)
our_kbl(surface_areas_yr_total)
```

```{r}
surface_areas_yr_total %>% 
  ggplot(aes(
    x = year_lsvi,
    y = surface_area,
    label = surface_area
  )) +
  geom_col(fill = "white") +
  geom_text(size = 3, vjust = 2) +
  scale_x_continuous(breaks = seq(2024, 2047, 1)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.4))
```

## Writing to GeoPackage
