---
title: "Designing LSVI-mapping in effect monitoring for the Programmatic Approach Nitrogen (PAN)"
date: today
published-title: Date
# csl: "`r if (curl::has_internet()) 'https://raw.githubusercontent.com/citation-style-language/styles/master/research-institute-for-nature-and-forest.csl' else system.file('research-institute-for-nature-and-forest.csl', package = 'INBOmd')`"
# bibliography: ../../../refs_mne_design.yaml
params:
  lang: nl
engine: knitr
format:
  html:
    embed-resources: true
    include-in-header: header_qmd.html
    theme: cosmo
    keep-md: true
    number-sections: true
    code-fold: true
    fig-width: 10
    fig-height: 5
    df-print: paged
    toc: true
    toc-location: left
    toc-depth: 4
    toc-expand: true
---

```{r setup, message = FALSE}
options(scipen = 999)
conflictRules("git2rdata", exclude = c("pull", "meta"))
conflictRules("tidyr", exclude = "extract")
conflictRules("knitr", exclude = "spin")
conflictRules("kableExtra", exclude = "group_rows")
lang <- params$lang
library(rprojroot)
projroot <- find_root(is_rstudio_project)
gitroot <- find_root(is_git_root)
datapath <- file.path(projroot, "data")
for (subdir in c(
  "binary/results",
  "text/raw"
)) {
  path <- file.path(datapath, subdir)
  if (!dir.exists(path)) dir.create(path, recursive = TRUE)
}
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(git2rdata)
library(sf)
library(terra)
library(n2khab)
source(file.path(projroot, "R/functions.R"))
opts_chunk$set(
  echo = TRUE,
  dpi = 300,
  fig.align = "center"
)
our_kbl <- function(df) {
  df %>% 
    kbl(table.attr = 'data-quarto-disable-processing="true"') %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "striped")
}
our_scroll_kbl <- function(df, width = NULL, height = NULL) {
  our_kbl(df) %>% 
    scroll_box(width = width, height = height)
}
```

## Settings and basic objects

Defining the regular expression for forest types.

```{r}
forest_types_regex <- "^9|^2180|^rbbppm"
```

Setting the theoretical first year of the first monitoring cycle as defined for reporting, i.e. taking into account the required alignment of monitoring cycles with reporting years.

```{r}
year_origin <- 2024L
```

Reading the reference checksum values.

```{r}
n2khab_data_checksums_reference <-
  read_vc(
    file = "n2khab_data_checksums_reference",
    root = file.path(datapath, "text/raw")
  ) %>%
  as_tibble()
```

Defining the data source versions to be used in this workflow.

```{r}
versions_required <- c(
  sac = "sac_2013-01-18",
  habitatmap_terr = "habitatmap_terr_2025_v99_interim",
  GRTSmh_diffres = "GRTSmh_diffres_v1"
)
stopifnot(all(
  versions_required %in% unique(n2khab_data_checksums_reference$version_id)
))
```

Checking the existence of the correct files in the correct directories.

```{r}
verify_n2khab_data(n2khab_data_checksums_reference, versions_required)
```

Defining domains, including their geometry.

```{r}
sac <- read_admin_areas(dsn = "sac")
sac_subset_5 <- c(
  "BE2100015", # Kalmthoutse Heide
  "BE2100024", # Vennen, heiden en moerassen rond Turnhout
  "BE2200028", # De Maten
  "BE2200035", # Mechelse heide en vallei van de Ziepbeek
  "BE2200039" # Voerstreek
)
for (sac_i in sac_subset_5) {
  sac %>%
    filter(sac_code == sac_i) %>%
    st_combine() %>%
    .[[1]] %>%
    assign(str_c("sfg_", tolower(sac_i)), ., envir = .GlobalEnv)
}
domains <- tribble(
  ~domain, ~domain_name_nl, ~domain_name_en, ~geometry,
  "BE2100015", "SBZ-H BE2100015", "SAC BE2100015", sfg_be2100015,
  "BE2100024", "SBZ-H BE2100024", "SAC BE2100024", sfg_be2100024,
  "BE2200028", "SBZ-H BE2200028", "SAC BE2200028", sfg_be2200028,
  "BE2200035", "SBZ-H BE2200035", "SAC BE2200035", sfg_be2200035,
  "BE2200039", "SBZ-H BE2200039", "SAC BE2200039", sfg_be2200039
) %>%
  mutate(across(where(is.character), fct)) %>%
  st_as_sf(crs = 31370)
domains
```

Reading spatial data of terrestrial types.

```{r}
hmt <- read_habitatmap_terr(
  keep_aq_types = FALSE,
  drop_7220 = FALSE,
  version = versions_required["habitatmap_terr"]
)
```

Reading GRTS raster cells at two specific resolutions.

```{r}
main_level <- 4
cells <- read_GRTSmh_diffres(level = main_level)
subcells <- read_GRTSmh_diffres(level = main_level - 1)
```

## Subsetting objects

Only polygons with a terrestrial _habitat_ type are retained.
We also mark which polygons contain at least one 'open type' (as opposed to 'forest type').

```{r}
hmt_occ_hab <- 
  hmt$habitatmap_terr_types %>%
  filter(!str_detect(type, "^rbb")) %>% 
  mutate(
    polygon_has_open_type = !all(str_detect(type, forest_types_regex)),
    .by = polygon_id
  )
hmt_pol_hab <- 
  hmt$habitatmap_terr_polygons %>% 
  inner_join(
    hmt_occ_hab %>%
      distinct(
        polygon_id,
        has_open_type = polygon_has_open_type
      ),
    join_by(polygon_id),
    relationship = "one-to-one",
    unmatched = c("drop", "error")
  ) %>% 
  select(polygon_id, description_orig, has_open_type)
```

Spatially subsetting habitat polygons according to domains (five SACs with custom measures).

```{r}
hmt_pol_hab_sacsubset <- hmt_pol_hab[domains, ]
```

Spatially subsetting cells and subcells according to domains.
We also include the cell address in the subcells object.

```{r}
cells_sacsubset <-
  cells[vect(domains), , drop = FALSE] %>% 
  as.polygons(round = FALSE, aggregate = FALSE) %>% 
  st_as_sf() %>% 
  as_tibble() %>% 
  rename(cell_address = 1) %>% 
  mutate(cell_address = as.integer(cell_address)) %>% 
  st_as_sf()
subcells_sacsubset <-
  subcells[vect(domains), , drop = FALSE] %>% 
  as.polygons(round = FALSE, aggregate = FALSE) %>% 
  st_as_sf() %>% 
  as_tibble() %>% 
  rename(subcell_address = 1) %>% 
  mutate(subcell_address = as.integer(subcell_address)) %>% 
  st_as_sf() %>% 
  st_join(cells_sacsubset, join = st_within, left = FALSE) %>% 
  relocate(geometry, .after = last_col())
```

Visualizing domains, cells and subcells.

```{r}
mvmap <- 
  mapview::mapview(domains, map.types = "CartoDB.Positron") +
  cells_sacsubset +
  subcells_sacsubset
mvmap@map %>% 
  inbospatial::add_wms_be_cartoweb("topo_grey")
```

## Habitat polygons with (sub)cell attribute

To join subcell and cell addresses to the habitat polygons, we need the centroids of the habitat polygons.

```{r}
hmt_pol_hab_sacsubset_centroid <- 
  st_intersection(
    hmt_pol_hab %>% st_sf(agr = "identity"),
    domains %>% st_geometry()
  ) %>% 
  st_sf(agr = "identity") %>% 
  st_centroid()
```

```{r}
#| eval: false
#| include: false
cells_sacsubset_hmt <-
  cells_sacsubset[hmt_pol_hab_sacsubset_centroid, ]
# all subcells belonging to cells_sacsubset_hmt
subcells_sacsubset_hmt <-
  subcells_sacsubset %>% 
  semi_join(
    cells_sacsubset_hmt %>% st_drop_geometry(),
    join_by(cell_address)
  )
```

```{r}
#| eval: false
#| include: false
mapview::mapview(domains) + cells_sacsubset_hmt + subcells_sacsubset_hmt + hmt_pol_hab_sacsubset_centroid
```

Joining subcell and cell addresses to the habitat polygons.

```{r}
hmt_pol_hab_sacsubset_subcell <- 
  hmt_pol_hab_sacsubset_centroid %>% 
  st_join(subcells_sacsubset, left = FALSE) %>% 
  st_drop_geometry() %>% 
  inner_join(
    hmt_pol_hab_sacsubset,
    .,
    join_by(polygon_id, description_orig, has_open_type),
    relationship = "one-to-one",
    unmatched = "error"
  ) %>% 
  relocate(geom, .after = last_col())
```

Inspect result:

```{r}
#| echo: false
#| paged.print: false
hmt_pol_hab_sacsubset_subcell
```


<!-- Plan: -->

<!-- distribute cells over 6 panels for open types -->
<!-- distribute (member) subcells over 12 (sub)panels for forest types -->

<!-- assign panels to years -->

<!-- generate 3 layers with the year attribute: cells, subcells, polygons -->

## Writing to GeoPackage
