---
title: "Designing LSVI-mapping in effect monitoring for the Programmatic Nitrogen Approach (PAS)"
date: today
author: "Floris Vanderhaeghe \u25c6 review (p.p.): Karen Wuyts, Ward Tamsyn"
published-title: Date
csl: "`r if (curl::has_internet()) 'https://raw.githubusercontent.com/citation-style-language/styles/master/research-institute-for-nature-and-forest.csl' else system.file('research-institute-for-nature-and-forest.csl', package = 'INBOmd')`"
bibliography: references.yaml
params:
  lang: nl
  rewrite_subcellpanels: FALSE
engine: knitr
format:
  html:
    embed-resources: true
    include-in-header: header_qmd.html
    theme: cosmo
    keep-md: true
    number-sections: true
    code-fold: true
    fig-width: 10
    fig-height: 5
    df-print: paged
    toc: true
    toc-location: left
    toc-depth: 4
    toc-expand: true
---

_Note: the source code of this document is available in a [GitHub repository](https://github.com/inbo/n2khab-lsvi-mapping-pas)._

```{r setup, message = FALSE, output = FALSE}
renv::restore()
options(scipen = 999)
conflictRules("git2rdata", exclude = c("pull", "meta"))
conflictRules("janitor", exclude = c("crosstab", "chisq.test", "fisher.test"))
conflictRules("tidyr", exclude = "extract")
conflictRules("knitr", exclude = "spin")
conflictRules("kableExtra", exclude = "group_rows")
lang <- params$lang
library(rprojroot)
projroot <- find_root(is_rstudio_project)
gitroot <- find_root(is_git_root)
datapath <- file.path(projroot, "data")
for (subdir in c(
  "binary/results",
  "text/raw",
  "text/results"
)) {
  path <- file.path(datapath, subdir)
  if (!dir.exists(path)) dir.create(path, recursive = TRUE)
}
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(janitor)
library(git2rdata)
library(sf)
library(units)
library(terra)
library(n2khab)
library(ggplot2)
library(mapview)
source(file.path(projroot, "R/functions.R"))
opts_chunk$set(
  echo = TRUE,
  dpi = 300,
  fig.align = "center"
)
our_kbl <- function(df) {
  df %>% 
    kbl(table.attr = 'data-quarto-disable-processing="true"') %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "striped")
}
our_scroll_kbl <- function(df, width = NULL, height = NULL) {
  our_kbl(df) %>% 
    scroll_box(width = width, height = height)
}
```

## Introduction

### LSVI in PAS effect monitoring

In PAS effect monitoring, three ways of data collection will be combined to make inferences about specific nitrogen-related indicators that are part of LSVI evaluation (biotic habitat quality):

- existing MHQ monitoring schemes at the Flemish level, which apply a sampling approach (covering most habitat types);
- LSVI-mapping of all **terrestrial** habitat types in the _five Special Areas of Conservation (SACs) with custom measures_, as a collaboration between PAS effect monitoring and LSVI-mapping efforts for the Nature Restoration Law (NRL). Here, LSVI-mapping will be applied to get accurate estimates at a relatively low cost;
- densification of the MHQ samples of **aquatic** habitat types in the _five SACs with custom measures_.
This is an effort specific to PAS effect monitoring.

Inferences about these LSVI indicators are needed at the type level, both at the level of Flanders and at the level of the individual SACS with custom measures.

This document concentrates on the **LSVI-mapping**.

### Requested LSVI-mapping

In LSVI-mapping one wants to determine LSVI (and its constituting indicators) for the terrestrial habitat types in all polygons where they occur, within a certain area and at a certain time (year).

In **PAS effect monitoring**, LSVI-mapping is needed for all terrestrial habitat types in the five Special Areas of Conservation (SACs) with custom measures, hereafter called the **domains**.

The revisit cycle is 6 years for all types except forest types, where it is 12 years.
This follows the monitoring cycles for MHQ and NRL at the Flemish level.

### Aim of this report

In this technical report we aim at designing the fieldwork of LSVI-mapping in PAS effect monitoring at a year-by-polygon resolution.
As a result, the necessary calendars and spatial layers must be generated that are needed for fieldwork planning.

More specifically, the design should meet several characteristics:

- Revisit cycles of 6 and 12 years in a serially alternating pattern.
This means that every year another panel (set of locations) is measured; the cycle restarts after all panels have been measured.
- Polygons must be grouped as (large enough) spatial clusters to achieve better fieldwork efficiency.
However to limit confounding between space (year) and time (location), the spatial clusters of each panel (year) should be:
  - well spread within and between the domains, meaning that they must not be too large;
  - selected randomly so that a single panel or a smaller set of panels can already be used for estimation according to a (stratified) cluster random sampling design.
  
  So a balance must be found between fieldwork efficiency and spread.
- The yearly surface area to be measured should (preferably) be similar between years.
- Spatial synergy between the revisit cycles of 6 years (open types) and 12 years (forest types), i.e. ideally the forest clusters are spatially nested in or very near to the open-type clusters of the same year.

### Methods

#### Hierarchical GRTS raster cells for polygon clustering

We need a spatial clustering method that allows to randomly select well-spread clusters per panel, with the clusters being spatially aligned between forest polygons and polygons with open types.
Therefore we choose relatively large raster cells to group polygons (cell assignment of a polygon being based on polygon centroid) and use GRTS addresses of these raster cells to assign cells (i.e. the polygons belonging to them) to panels.
From here on, we use the term 'cell' to refer to the object that defines a polygon cluster, since the cell is the technical implementation of the clustering.

We will use the raster layers of (reverse) GRTS addresses in the data source `GRTSmh_diffres` [@vanderhaeghe_grtsmh_diffres:_2019], which is based on the `GRTSmaster_habitats` data source [@onkelinx_thierry_2019_2682323].
Only the cells that overlap the domains are used.
The (reverse) GRTS addresses allow drawing spatially well-spread random samples, hence they equally allow to combine spatial spread and randomization in the panel memberships of cells (hence of the polygons assigned to the cells).

The data source `GRTSmh_diffres` contains the GRTS cell hierarchy at different resolutions.
In order to spatially align the 6 panels of the 6-year cycle (polygons with an open type) with the 12 panels of the 12-year cycle (forest-only polygons), the latter will be nested inside the former by using two consecutive cell resolutions of the GRTS hierarchy.
We call the cells for the forest-only polygons _subcells_; the corresponding panels are _subpanels_.
More specifically, each panel (6-year) contains 2 subpanels.
For the panels applied to polygons with an open type, we just speak of _cells_.

#### Panel membership design

Cells are assigned to panels based on a domain-specific partitioning of the cell series ordered by GRTS address.
A partitioning is chosen that tries to achieve an equal number of cells per panel.

Cells contain 1 to 4 subcells: less than 4 subcells may occur on the domain boundaries, i.e. as a consequence of the intersection.
To spatially align panel memberships with subpanel memberships, the subcells of a cell are randomly distributed among the two subpanels of the cell's panel, making use of the subcells' GRTS addresses and applying a constraint of maximum 1 subcell difference for this division, i.e. equal partitioning is used if there are 2 or 4 subcells.

By directly assigning cells to panels and subcells to subpanels and by storing these panel memberships, we obtain a method of clustering polygons that remains applicable over time.
This is because the (sub)cells cover the entire domains, so that they remain useful regardless of changes in habitat polygons over time.
Only if the set of domains is expanded or shrinked, the panel memberships of (sub)cells must be updated (expanded or shrinked).
However the panel memberships within a domain will remain stable as long as the domain geometry is stable (which is expected).

#### Spatial layer of terrestrial habitat types

We will use the latest version of the `habitatmap_terr` data source -- either the regular or the interim version [@wouters_interim_2025; @wouters_habitatmap_terr_2026] -- to:

- define the polygons that are to be allocated to (sub)panels;
- distinguish polygons with an open type from forest-only polygons;
- calculate total surface areas for LSVI measurements, by domain, year and polygon category.

In fact, using a static data source for these purposes will always run behind the current state of the dynamic process of field mapping, where new polygons are created and existing ones may change or disappear.
This also means that the (sub)cell panel memberships are actually more useful than polygon panel memberships, since they are applicable at any time to determine (sub)panel membership of a polygon.
It depends on the polygon category (does it contain an open type?) **whether the cells or the subcells must be used to determine the polygon's (sub)panel**.

#### Revisit design

The revisit design is a serially alternating pattern and it can be operationally defined by assigning years to panels.
For that, we take two cycles of the revisit design with the longest cycle duration (for forest polygons), so we assign years over a period of 2 * 12 y = 24 y.

Next, this is applied to the panel memberships of cells, subcells and polygons to achieve a revisit calendar of cells, subcells and polygons.

#### Evaluation

Several aspects of the resulting calendars are explored to allow further evaluation or optimization:

- total number of (sub)cells per (sub)panel;
- temporal evenness of yearly polygon area per type;
- temporal evenness of total yearly polygon area, also per polygon category.

Apart from these results, the spatial layers have also been explored in a GIS environment, especially to assess the trade-off between cell size, spatial spread and fieldwork feasability.
Changes are motivated in git commit messages.

The most desirable outcome is a combination of overall temporal evenness of efforts, sufficiently small polygon clusters (allowing better spatial spread per year) and an acceptable fieldwork feasability.

### Sidenotes about inference

While (design-based) inference methods are not part of this document, they are tied to the design so it's useful to already list different aspects of inference.

The hierarchy of years, spatially well-spread clusters and polygons leads to a compound spatiotemporal sampling design that has _cluster sampling_ at its center [@cochran_sampling_1977; @de_gruijter_sampling_2006].
The clusters are all polygons (of the same category) that belong to the same cell or subcell (cell/subcell depends on the polygon category), and they are sampled using _GRTS_.

It's spatiotemporal _sampling_, not a census, even while all locations are visited after 6 or 12 years.
A census in the spatiotemporal sense means visiting all locations every year (pure panel design), while in this case every year only a sample of locations is measured.

Inference about **status** (design-based) refers to a single cycle or a part thereof (subset of years).
Since we don't _sample_ years but take a spatial sample _every_ year, we have temporal stratification by year.
So we end up with a **stratified GRTS-cluster sampling design**.

Inference about **trend** can involve estimation of change (difference) between two consecutive cycles, which can be done design-based, or the estimation of a  model trend, which can be done in a hybrid model- & design-based approach.

- difference between two cycles: this needs two cycles or at least two repetitions of a year subset of the cycles, which is possible already before the second cycle is completed.
This is best done (more efficient, more power) by spatially aggregating the location-wise (temporal) differences between two cycles.
The involved pairs of years can serve as strata.
- modelled trend: this is most robust after three cycles, as the estimation can then be done per location (e.g. a location-specific trend prediction by a spatiotemporal model that uses all locations) followed by design-based spatial aggregation of the local trend parameters.
It is also possible (but much less powerful) to fit a trend after a shorter time by modelling the trend of the yearly (spatially aggregated) status.

In both cases, the location-specific approaches come down to applying the (design-based) inference for the GRTS-cluster design after temporal aggregation (design- or model-based) at location level.

Inference will be done for each habitat main type or subtype; perhaps also for some grouping of habitat (sub)types.

Even though type units are not directly sampled -- only polygons are -- and even though a polygon often contains more than one habitat type, the cell-based cluster approach can be regarded as being insensitive to this, because all habitat polygons in a cluster are sampled anyway.
For inference will want to take into account each polygon's surface area and the proportion covered by the habitat (phab) inside each polygon in order to define 'cluster size' (all units of a cluster), as expressed by the type's surface area in a cluster.

Further, we must be aware that forest types occur both in forest-only polygons and in polygons with an open type.
For these types, appropriate weighting is needed to take into account that we assess them in forest-only polygons at half the speed as the polygons with open types, so forest-type measurements from the latter will are overrepresented in the data.

The differences in cluster sizes are typically accounted for in inference for cluster random sampling [@cochran_sampling_1977].


## Settings and basic objects

Defining the regular expression for forest types.

```{r}
#| code-fold: show
forest_types_regex <- "^9|^2180|^rbbppm"
```

Setting the theoretical first year of the first monitoring cycle as defined for reporting, i.e. taking into account the required alignment of monitoring cycles with reporting years.
Also, set the cycle duration for the open types.

```{r}
#| code-fold: show
year_origin <- 2024L
cycle_opentypes <- 6L
```

Reading the reference checksum values.

```{r}
n2khab_data_checksums_reference <-
  read_vc(
    file = "n2khab_data_checksums_reference",
    root = file.path(datapath, "text/raw")
  ) %>%
  as_tibble()
```

Defining the data source versions to be used in this workflow.

```{r}
#| code-fold: show
versions_required <- c(
  sac = "sac_2013-01-18",
  habitatmap_terr = "habitatmap_terr_2025_v99_interim",
  GRTSmh_diffres = "GRTSmh_diffres_v1"
)
stopifnot(all(
  versions_required %in% unique(n2khab_data_checksums_reference$version_id)
))
```

Checking the existence of the correct files in the correct directories.

```{r}
verify_n2khab_data(n2khab_data_checksums_reference, versions_required)
```

Defining domains, including their geometry.

```{r}
sac <- read_admin_areas(dsn = "sac")
sac_subset_5 <- c(
  "BE2100015", # Kalmthoutse Heide
  "BE2100024", # Vennen, heiden en moerassen rond Turnhout
  "BE2200028", # De Maten
  "BE2200035", # Mechelse heide en vallei van de Ziepbeek
  "BE2200039" # Voerstreek
)
for (sac_i in sac_subset_5) {
  sac %>%
    filter(sac_code == sac_i) %>%
    st_combine() %>%
    .[[1]] %>%
    assign(str_c("sfg_", tolower(sac_i)), ., envir = .GlobalEnv)
}
domains <- tribble(
  ~domain, ~domain_name_nl, ~domain_name_en, ~geometry,
  "BE2100015", "SBZ-H BE2100015", "SAC BE2100015", sfg_be2100015,
  "BE2100024", "SBZ-H BE2100024", "SAC BE2100024", sfg_be2100024,
  "BE2200028", "SBZ-H BE2200028", "SAC BE2200028", sfg_be2200028,
  "BE2200035", "SBZ-H BE2200035", "SAC BE2200035", sfg_be2200035,
  "BE2200039", "SBZ-H BE2200039", "SAC BE2200039", sfg_be2200039
) %>%
  mutate(across(where(is.character), fct)) %>%
  st_as_sf(crs = 31370)
domains
```

Reading spatial data of terrestrial types.

```{r}
#| code-fold: show
hmt <- read_habitatmap_terr(
  keep_aq_types = FALSE,
  drop_7220 = FALSE,
  version = versions_required["habitatmap_terr"]
)
```

Reading GRTS raster cells at two specific resolutions.

```{r}
#| code-fold: show
main_level <- 4
cells <- read_GRTSmh_diffres(level = main_level)
subcells <- read_GRTSmh_diffres(level = main_level - 1)
```

## Subsetting objects

Only polygons with a terrestrial _habitat_ type are retained.
We also mark which polygons contain at least one 'open type' (as opposed to 'forest type').

```{r}
hmt_occ_hab <- 
  hmt$habitatmap_terr_types %>%
  filter(!str_detect(type, "^rbb")) %>% 
  mutate(
    polygon_has_open_type = !all(str_detect(type, forest_types_regex)),
    .by = polygon_id
  )
hmt_pol_hab <- 
  hmt$habitatmap_terr_polygons %>% 
  inner_join(
    hmt_occ_hab %>%
      distinct(
        polygon_id,
        has_open_type = polygon_has_open_type
      ),
    join_by(polygon_id),
    relationship = "one-to-one",
    unmatched = c("drop", "error")
  ) %>% 
  select(polygon_id, description_orig, has_open_type)
```

Spatially subsetting habitat polygons according to domains (five SACs with custom measures), i.e. all habitat polygons are retained that intersect the domains.

```{r}
hmt_pol_hab_sacsubset <- hmt_pol_hab[domains, ]
```

Inspecting result:

```{r}
#| paged.print: false
#| code-fold: show
hmt_pol_hab_sacsubset
```

```{r}
#| code-fold: show
table(hmt_pol_hab_sacsubset$has_open_type)
```

Spatially subsetting cells and subcells according to domains.
The (sub)cells are retained when they overlap the domain (the cell center does not have to overlap the domain).
We also include the cell address in the subcells object.

```{r}
cells_sacsubset_pre <-
  extract(cells, domains, ID = FALSE, touches = TRUE, cells = TRUE)$cell %>% 
  cells[., drop = FALSE] %>% 
  as.polygons(round = FALSE, aggregate = FALSE) %>% 
  st_as_sf() %>% 
  as_tibble() %>% 
  rename(cell_address = 1) %>% 
  mutate(cell_address = as.integer(cell_address)) %>% 
  st_as_sf()
# adding domain attribute
cells_sacsubset <- 
  cells_sacsubset_pre %>% 
  st_join(
    domains %>% select(domain),
    join = st_intersects,
    left = FALSE
  ) %>% 
  relocate(geometry, .after = last_col()) %>% 
  arrange(domain, cell_address)
# we assume that no cells are member of two domains at the same time (if not
# fulfilled we need to re-evaluate how we cope with domain-wise cell panel
# membership)
stopifnot(nrow(cells_sacsubset_pre) == nrow(cells_sacsubset))
subcells_sacsubset <-
  extract(subcells, domains, ID = FALSE, touches = TRUE, cells = TRUE)$cell %>% 
  subcells[., drop = FALSE] %>% 
  as.polygons(round = FALSE, aggregate = FALSE) %>% 
  st_as_sf() %>% 
  as_tibble() %>% 
  rename(subcell_address = 1) %>% 
  mutate(subcell_address = as.integer(subcell_address)) %>% 
  st_as_sf() %>% 
  st_join(cells_sacsubset, join = st_within, left = FALSE) %>% 
  arrange(cell_address, subcell_address) %>% 
  relocate(geometry, .after = last_col())
```

Inspecting the results:

```{r}
#| paged.print: false
#| code-fold: show
cells_sacsubset
```

```{r}
#| paged.print: false
#| code-fold: show
subcells_sacsubset
```

Visualizing domains, cells and subcells.

```{r}
#| fig-height: 8
#| fig-width: 8
mvmap <- 
  mapview(
    subcells_sacsubset,
    map.types = "CartoDB.Positron",
    col.regions = "blue",
    label = NA
  ) +
  mapview(
    cells_sacsubset,
    col.regions = "red",
    lwd = 1.5,
    alpha.regions = 0.5,
    label = NA
  ) +
  mapview(
    domains,
    alpha.regions = 0,
    col.regions = "pink",
    color = "pink",
    lwd = 2,
    label = NA
  )
mvmap@map %>% 
  inbospatial::add_wms_be_cartoweb("topo_grey")
```

## Habitat polygons with (sub)cell attribute

To join subcell and cell addresses to the habitat polygons in a one-to-one way, we will use the centroids of the habitat polygons.

First creating these centroids.

```{r}
hmt_pol_hab_sacsubset_centroid <- 
  st_intersection(
    hmt_pol_hab_sacsubset %>% st_sf(agr = "identity"),
    domains %>% st_geometry()
  ) %>% 
  st_sf(agr = "identity") %>% 
  st_centroid()
```

```{r}
stopifnot(nrow(hmt_pol_hab_sacsubset) == nrow(hmt_pol_hab_sacsubset_centroid))
```

```{r}
#| include: false
if (FALSE) {
  cells_sacsubset_hmt <-
    cells_sacsubset[hmt_pol_hab_sacsubset_centroid, ]
  # all subcells belonging to cells_sacsubset_hmt
  subcells_sacsubset_hmt <-
    subcells_sacsubset %>% 
    semi_join(
      cells_sacsubset_hmt %>% st_drop_geometry(),
      join_by(cell_address)
    )
}
```

```{r}
#| include: false
if (FALSE) {
  mapview::mapview(domains) +
    cells_sacsubset_hmt +
    subcells_sacsubset_hmt +
    hmt_pol_hab_sacsubset_centroid
}
```

Joining subcell and cell addresses to the habitat polygons.

```{r}
hmt_pol_hab_sacsubset_subcell <- 
  hmt_pol_hab_sacsubset_centroid %>% 
  st_join(subcells_sacsubset, left = FALSE) %>% 
  st_drop_geometry() %>% 
  inner_join(
    hmt_pol_hab_sacsubset,
    .,
    join_by(polygon_id, description_orig, has_open_type),
    relationship = "one-to-one",
    unmatched = "error"
  ) %>% 
  relocate(geom, .after = last_col())
```

Inspecting result:

```{r}
#| paged.print: false
#| code-fold: show
hmt_pol_hab_sacsubset_subcell
```


## Implement membership design

The panel membership will be defined at the level of cells and subcells, using their (reverse) GRTS address order.

For panel memberships, we start from the `cells_sacsubset` and `subcells_sacsubset` objects without taking into account habitat polygon distribution, to make these membership more stable through time, i.e. not sensitive to changes in the habitatmap.
This approach is expected to still lead to a relatively even spread of the habitat polygons over panels, which we will check.

Note that for subcells we use _subpanels_, i.e. a numbering to be interpreted within the cell panel.

We always recalculate the cell panel membership.
If the domains ever change, this may cause a few shifts but these should still be relatively small given the large number of cells compared to the number of panels.
Still, we will protect against effects of domain additions or removals by calculating panel membership domain-wise, not across domains.

```{r}
set.seed(20260908) # not clear whether this may have effect
cells_sacsubset_panel <- 
  cells_sacsubset %>% 
  st_drop_geometry() %>% 
  mutate(
    panel_cell = cell_address %>% 
      cut_number(n = cycle_opentypes) %>% 
      as.integer(),
    .by = domain
  )
```

Distribution of cells over panels:

```{r}
#| code-fold: show
table(cells_sacsubset_panel$panel_cell)
```

We have calculated subpanel membership of subcells in the past and saved it on disk.
We recalculate the subpanel membership (and optionally rewrite it) to test for equality with the version on disk.

```{r}
set.seed(20260908)
subcells_sacsubset_subpanel_recalc <- 
  subcells_sacsubset %>% 
  st_drop_geometry() %>% 
  inner_join(
    cells_sacsubset_panel,
    join_by(cell_address, domain),
    relationship = "many-to-one",
    unmatched = "error"
  ) %>% 
  mutate(
    subpanel_subcell = divide_subcell_addresses_in_two(subcell_address),
    .by = cell_address
  )
```

```{r}
if (
  params$rewrite_subcellpanels ||
  !file.exists(file.path(
    datapath,
    "text/results/subcells_sacsubset_subpanel.yml"
  ))
) {
  subcells_sacsubset_subpanel_recalc %>% 
    write_vc(
      file = "subcells_sacsubset_subpanel",
      root = file.path(datapath, "text/results"),
      digits = 15,
      optimize = TRUE,
      strict = FALSE,
      sorting = c("cell_address", "subcell_address")
    ) %>% 
    convert_line_endings(root = file.path(datapath, "text/results"))
}
```

Using the subcell subpanel membership that has been saved before, and test that it is the same.

```{r}
subcells_sacsubset_subpanel <- 
  read_vc(
    file = "subcells_sacsubset_subpanel",
    root = file.path(datapath, "text/results")
  ) %>% 
  as_tibble()
# if this test fails, manual intervention is needed to decide on changes
stopifnot(all.equal(
  subcells_sacsubset_subpanel,
  subcells_sacsubset_subpanel_recalc,
  check.attributes = FALSE
))
```

## Define revisit design

We define the revisit design as an assignment of calendar years to panels and subpanels for a period of 24 years.

The revisit design is different between polygons that have at least one open type and forest polygons:

- when open types are present, a cycle duration of 6 years is used in a serially alternating pattern [1-5];
- when a polygon is forest-only, a cycle duration of 12 years is used in a serially alternating pattern [1-11], implemented by creating 2 _subpanels_ within each panel for open types.
In this way the panel membership of open types is taken into account.

Generating the panel calendar.

```{r}
panel_calendar <- 
  subcells_sacsubset_subpanel %>% 
  expand(
    panel_cell = full_seq(panel_cell, 1L),
    subpanel_subcell = full_seq(subpanel_subcell, 1L)
  ) %>% 
  arrange(subpanel_subcell, panel_cell) %>% 
  bind_rows(., .) %>% 
  bind_cols(year_lsvi = year_origin + seq_len(nrow(.)) - 1L)
```

Inspecting the result:

```{r}
#| paged.print: false
#| code-fold: show
panel_calendar %>% print(n = Inf)
```

## Generate calendars for cells, subcells and polygons

We generate the cell calendar (open types), the subcell calendar (forest types) and polygon calendar by joining the panel years to the (sub)panels.
Only in the polygon calendar it is possible to integrate both revisit designs without spatial overlaps.

```{r}
cells_calendar <- 
  cells_sacsubset_panel %>% 
  inner_join(
    panel_calendar %>% 
      select(panel_cell, year_lsvi),
    join_by(panel_cell),
    relationship = "many-to-many",
    unmatched = "error"
  )
```

Structure of the cell calendar (open types):

```{r}
#| code-fold: show
glimpse(cells_calendar)
```

```{r}
subcells_calendar <- 
  subcells_sacsubset_subpanel %>% 
  inner_join(
    panel_calendar,
    join_by(panel_cell, subpanel_subcell),
    relationship = "many-to-many",
    unmatched = "error"
  )
```

Structure of the subcell calendar (forest types):

```{r}
#| code-fold: show
glimpse(subcells_calendar)
```

In the polygon calendar, we set the subcell address and subcell subpanel as missing if the polygon has at least one open type, to avoid confusion: subcells are not relevant in the revisit design of those polygons.

```{r}
polygons_calendar <- 
  hmt_pol_hab_sacsubset_subcell %>% 
  st_drop_geometry() %>% 
  inner_join(
    cells_calendar,
    join_by(cell_address, domain),
    relationship = "many-to-many",
    unmatched = c("error", "drop")
  ) %>% 
  left_join(
    subcells_calendar,
    join_by(
      subcell_address,
      cell_address,
      domain,
      panel_cell,
      # matching years: (forest) subcell revisit years match (open) cell revisit
      # years
      year_lsvi
    ),
    # multiple polygons can have their centroid in the same subcell
    relationship = "many-to-one",
    unmatched = "drop"
  ) %>% 
  # keep subcell_address and subpanel_subcell only for forest-only polygons
  mutate(across(
    c(subcell_address, subpanel_subcell),
    \(x) ifelse(has_open_type, NA_integer_, x)
  )) %>% 
  # drop the empty subcell revisits in case of forest-only polygons
  filter_out(!has_open_type & is.na(subpanel_subcell)) %>% 
  relocate(
    c(
      year_lsvi,
      domain,
      cell_address,
      panel_cell,
      subcell_address,
      subpanel_subcell
    ),
    .after = has_open_type
  ) %>% 
  arrange(year_lsvi, domain, cell_address, subcell_address)
```

Structure of the polygon calendar (all types):

```{r}
#| code-fold: show
glimpse(polygons_calendar)
```


## Derive applied, spatial objects

We also derive geospatial objects that flatten the years in a single row.
These objects are stripped from background information to avoid confusion in applications.

```{r}
opentypes_cellcalendar_lsvi_pas <-
  cells_sacsubset %>% 
  inner_join(
    cells_calendar %>% 
      summarise(
        years_lsvi = str_flatten(sort(year_lsvi), collapse = "|") %>% factor(),
        .by = c(cell_address, domain)
      ),
    join_by(cell_address, domain),
    relationship = "one-to-one",
    unmatched = "error"
  ) %>%
  relocate(geometry, .after = last_col()) %>% 
  arrange(domain, years_lsvi, cell_address)
```

Inspecting geospatial form of the cell calendar (open types):

```{r}
#| paged.print: false
#| code-fold: show
opentypes_cellcalendar_lsvi_pas
```

```{r}
foresttypes_subcellcalendar_lsvi_pas <-
  subcells_sacsubset %>% 
  select(-cell_address) %>% 
  inner_join(
    subcells_calendar %>% 
      summarise(
        years_lsvi = str_flatten(sort(year_lsvi), collapse = "|") %>% factor(),
        .by = c(subcell_address, domain)
      ),
    join_by(subcell_address, domain),
    relationship = "one-to-one",
    unmatched = "error"
  ) %>%
  relocate(geometry, .after = last_col()) %>% 
  arrange(domain, years_lsvi, subcell_address)
```

Inspecting geospatial form of the subcell calendar (forest types):

```{r}
#| paged.print: false
#| code-fold: show
foresttypes_subcellcalendar_lsvi_pas
```

```{r}
polygons_calendar_lsvi_pas <-
  hmt_pol_hab_sacsubset %>% 
  inner_join(
    polygons_calendar %>% 
      summarise(
        years_lsvi = str_flatten(sort(year_lsvi), collapse = "|") %>% factor(),
        .by = c(polygon_id, description_orig, has_open_type, domain)
      ),
    join_by(polygon_id, description_orig, has_open_type),
    relationship = "one-to-one",
    unmatched = "error"
  ) %>%
  relocate(geom, .after = last_col()) %>% 
  arrange(domain, years_lsvi, polygon_id)
```

Inspecting geospatial form of the polygon calendar (all types):

```{r}
#| paged.print: false
#| code-fold: show
polygons_calendar_lsvi_pas
```


## Explore results

```{r}
plot_category_areas <- function(df, response_var, fill = NULL) {
  df %>% 
  mutate({{response_var}} := round({{response_var}}, 2)) %>% 
  ggplot(aes(
    x = year_lsvi,
    y = {{response_var}},
    fill = {{fill}},
    label = {{response_var}}
  )) +
  geom_col() +
  scale_x_continuous(breaks = seq(
    year_origin,
    year_origin + 4 * cycle_opentypes - 1,
    1
  )) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.4),
    legend.position = "bottom"
  )
}
```

### Visualizing revisit design: total cell and subcell numbers per year

```{r}
cell_count_yr <- 
  cells_sacsubset %>% 
  select(cell_address) %>% 
  st_drop_geometry() %>% 
  inner_join(
    cells_calendar %>%
      select(domain, cell_address, year_lsvi),
    join_by(cell_address),
    relationship = "one-to-many",
    unmatched = "error"
  ) %>% 
  count(domain, year_lsvi, name = "number_of_cells")
```

Total number of cells per year:

```{r}
cell_count_yr %>% 
  summarise(number_of_cells = sum(number_of_cells), .by = year_lsvi) %>% 
  our_kbl()
```

```{r}
cell_count_yr %>% 
  plot_category_areas(response_var = number_of_cells, fill = domain) +
  ggtitle("Total number of cells intersecting the domains, irrespective of the presence of habitat")
```

```{r}
subcell_count_yr <- 
  subcells_sacsubset %>% 
  select(subcell_address) %>% 
  st_drop_geometry() %>% 
  inner_join(
    subcells_calendar %>%
      select(domain, subcell_address, year_lsvi),
    join_by(subcell_address),
    relationship = "one-to-many",
    unmatched = "error"
  ) %>% 
  count(domain, year_lsvi, name = "number_of_subcells")
```

Total number of subcells per year:

```{r}
subcell_count_yr %>% 
  summarise(number_of_subcells = sum(number_of_subcells), .by = year_lsvi) %>% 
  our_kbl()
```

```{r}
subcell_count_yr %>% 
  plot_category_areas(response_var = number_of_subcells, fill = domain) +
  ggtitle("Total number of subcells intersecting the domains, irrespective of the presence of habitat")
```

### Distribution of types by year

Spread of surface area by type and year:

```{r}
#| fig-height: 18
#| fig-width: 14
hmt_pol_hab_sacsubset %>% 
  select(polygon_id) %>% 
  mutate(polygon_area = st_area(.) %>% set_units("ha")) %>% 
  st_drop_geometry() %>% 
  inner_join(
    hmt_occ_hab %>% select(polygon_id, type, phab),
    join_by(polygon_id),
    relationship = "one-to-many",
    unmatched = c("error", "drop")
  ) %>% 
  mutate(
    type_area = polygon_area * phab / 100,
    .keep = "unused"
  ) %>% 
  mutate(is_forest_type = str_detect(type, forest_types_regex)) %>% 
  # dropping type presences with phab = 0 (not useful for type area)
  filter(type_area > set_units(0, "ha")) %>% 
  inner_join(
    polygons_calendar %>% 
      select(polygon_id, domain, year_lsvi),
    join_by(polygon_id),
    relationship = "many-to-many",
    # dropping calendar items where phab = 0
    unmatched = c("error", "drop")
  ) %>% 
  summarise(
    type_area = sum(type_area),
    .by = c(type, year_lsvi)
  ) %>% 
  ggplot(aes(
    x = year_lsvi,
    y = type_area,
    label = type_area
  )) +
  geom_col() +
  scale_x_continuous(breaks = seq(year_origin, year_origin + 23, 1)) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.4),
    legend.position = "bottom"
  ) +
  facet_wrap(~type, scales = "free_y", ncol = 5)
```


### Visualizing LSVI surface areas by year

Total areas according to `has_open_type` and panel (as represented by groups of years).
The table rows represent mutually exclusive polygon sets.

```{r}
polygons_calendar_lsvi_pas %>% 
  mutate(polygon_area = st_area(.) %>% set_units("ha")) %>% 
  st_drop_geometry() %>% 
  summarise(
    surface_area = sum(polygon_area) %>% round(2),
    .by = c(has_open_type, years_lsvi)
  ) %>% 
  arrange(desc(has_open_type), years_lsvi) %>% 
  our_kbl()
```

The same information, visualized by year:

```{r}
surface_areas_yr <- 
  hmt_pol_hab_sacsubset %>% 
  select(polygon_id) %>% 
  mutate(polygon_area = st_area(.) %>% set_units("ha")) %>% 
  st_drop_geometry() %>% 
  inner_join(
    polygons_calendar %>%
      select(domain, polygon_id, has_open_type, year_lsvi),
    join_by(polygon_id),
    relationship = "one-to-many",
    unmatched = "error"
  ) %>% 
  mutate(polygon_category = ifelse(
    has_open_type,
    "contains_opentype",
    "forest_only"
  )) %>% 
  summarise(
    surface_area = sum(polygon_area),
    .by = c(domain, polygon_category, year_lsvi)
  )
```

```{r}
surface_areas_yr %>% 
  summarise(
    surface_area = sum(surface_area),
    .by = c(polygon_category, year_lsvi)
  ) %>% 
  plot_category_areas(response_var = surface_area, fill = polygon_category) +
  geom_text(size = 3, vjust = 2)
```

Distinguishing domains:

```{r}
surface_areas_yr %>% 
  summarise(
    surface_area = sum(surface_area),
    .by = c(domain, polygon_category, year_lsvi)
  ) %>% 
  plot_category_areas(response_var = surface_area, fill = polygon_category) +
  facet_wrap(~domain, scales = "free_y")
```

Total areas by year:

```{r}
surface_areas_yr_total <- 
  surface_areas_yr %>% 
  summarise(
    surface_area = sum(surface_area),
    .by = c(year_lsvi)
  ) %>% 
  mutate(surface_area = round(surface_area, 2)) %>% 
  arrange(year_lsvi)
our_kbl(surface_areas_yr_total)
```

```{r}
surface_areas_yr_total %>% 
  ggplot(aes(
    x = year_lsvi,
    y = surface_area,
    label = surface_area
  )) +
  geom_col(fill = "white") +
  geom_text(size = 3, vjust = 2) +
  scale_x_continuous(breaks = seq(
    year_origin,
    year_origin + 4 * cycle_opentypes - 1,
    1
  )) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.4))
```

### Addendum: yearly cell area

We calculate and visualize the yearly surface area within domains that is covered by the cells, according to the cell panel calendar.
This is relevant because the cells will at the same time serve for mapping BWK and habitat types over the whole surface of each domain: LSVI-mapping happens in combination with BWK & habitat mapping.
We can use the revisit cycle of 6 years as approximation, hence use the entire cells (clipped by domain), even though in practice large forest polygons will only be mapped every 12 years.

```{r}
cell_areas_domain_year <- 
  opentypes_cellcalendar_lsvi_pas %>% 
  select(years_lsvi) %>% 
  st_sf(agr = "constant") %>% 
  st_intersection(
    domains %>% 
      select(domain) %>% 
      st_sf(agr = "identity")
  ) %>% 
  mutate(surface_area = st_area(.) %>% set_units("ha")) %>% 
  st_drop_geometry() %>% 
  summarize(
    surface_area = sum(surface_area),
    .by = c(years_lsvi, starts_with("domain"))
  )
cell_areas_domain_year %>% 
  mutate(surface_area = round(surface_area, 2)) %>% 
  pivot_wider(names_from = domain, values_from = surface_area) %>% 
  adorn_totals(where = "col", name = "total") %>% 
  mutate(total = set_units(total, "ha")) %>% 
  as_tibble() %>% 
  our_kbl()
```

```{r}
cell_areas_domain_year %>% 
  mutate(cycle_year = as.integer(years_lsvi)) %>% 
  ggplot(aes(x = cycle_year, y = surface_area, fill = domain)) +
  geom_col() +
  scale_x_continuous(breaks = seq_len(cycle_opentypes))
```



## Writing to GeoPackage

```{r}
gpkg_path <- file.path(datapath, "binary/results/lsvi_mapping_pas.gpkg")
polygons_calendar_lsvi_pas %>%
  write_sf(
    gpkg_path,
    layer = "polygons_calendar_lsvi_pas",
    delete_dsn = TRUE
  )
opentypes_cellcalendar_lsvi_pas %>%
  write_sf(
    gpkg_path,
    layer = "opentypes_cellcalendar_lsvi_pas",
    delete_layer = TRUE
  )
foresttypes_subcellcalendar_lsvi_pas %>%
  write_sf(
    gpkg_path,
    layer = "foresttypes_subcellcalendar_lsvi_pas",
    delete_layer = TRUE
  )
domains %>%
  write_sf(
    gpkg_path,
    layer = "sac_custom_measures_pas",
    delete_layer = TRUE
  )
```

Inspecting the file layers.

```{r}
#| paged.print: false
#| code-fold: show
st_layers(gpkg_path)
```


## References {-}

::: {#refs}
:::


## Appendix: used environment {-}

### Platform {-}

```{r session-info, results = "asis", echo=FALSE}
si <- sessioninfo::session_info()
p <- si$platform %>%
  do.call(what = "c")
if ("sf" %in% si$packages$package) {
  p <- c(p, sf::sf_extSoftVersion())
  p <- p[names(p) != "proj.4"]
}
if ("rgrass" %in% si$packages$package) {
  p <- c(p, GRASS = link2GI::findGRASS()[1, "version"])
}
sprintf("- **%s**: %s\n", names(p), p) %>%
  cat(sep = "")
```

### R packages {-}

```{r results = "asis", echo=FALSE}
#| label: tbl-sessioninfo
#| tbl-cap: "Loaded R packages"
si$packages %>%
  as_tibble() %>%
  select(package, loadedversion, date, source) %>%
  pander::pandoc.table(split.table = Inf)
```
